Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-03-11 ResponseLetter2.tex

\begin{document}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{Reviewer $\#$1}

Minor Edits:

1. Review all Figure titles (main/supplement) and turn the Figure titles into impact statements of the main result, like you did for Figures 3, 5, and 6, which give your reader an immediate connection to the purpose of the figure and the take-away. Change Figure 3 title to ``\ldots evolutionary rescue of tumors under drug treatment \ldots''. In Table 1, ``\ldots 0.01, instead \dots''. In Appendix A, ``\ldots (Kendall, 1948). This allows\ldots''. In Figure 5, ``Standing genetic variation will drive drug resistance when the sensitive cell population is rapidly declining.'' In Figure 5, ``\ldots $N_a^*$ is a function\ldots''

@@

\textcolor{blue}{We have changed the title of Figure 4 and added a title to Figure S5. Change S18}

2. Please do a search for ``very'' across the whole manuscript and remove. The world holds no meaning here. If there is a need to describe an overwhelming amount of anything, then quantify it.

@@

\textcolor{blue}{Change S1}

3. There is a tendency in the manuscript to use some kind of connection word to start sentences, and to do it back-to-back. For example, in Lines 22-25 in the Abstract, ``Additionally\ldots Finally,'' and lines 51-54 in the Introduction, ``Consequently\ldots Furthermore.'' It is a bit distracting and unnecessary and detracts from the overall quality of the paper, which is excellent. Go through the manuscript and remove as much of this is possible.

@@

\textcolor{blue}{Change S16}

4. Line 16, remove ``, e.g.,''

\textcolor{blue}{Change S2}

5. Line 25, ``\ldots small to intermediate size primary and secondary tumors.'' You use small and intermediate in the discussion. Not sure if you need primary and secondary here.

\textcolor{blue}{We decided to keep ``small secondary tumors'' as this is an important insight from our model.}

6. Global search on ``missegregation'' to make sure you use it with/without hyphen consistently.

@@

10. Line 221, put "Table 1" in parentheses in a previous sentence.

\textcolor{blue}{Change S7}

11. Sometimes you set variable names between commas, sometimes in parenthesis, sometimes without a leading comma. I would recommend a consistent strategy.

\textcolor{blue}{We believe that both parenthesis and commas are appropriate depending on the context.}

12. Line 297, ``\ldots as it reduced the threshold tumor sized needed for rescue.''

@@

\textcolor{blue}{Change S8}

13. Lines 334-343: ``In the above\ldots However\ldots Moreover\ldots Hence\dots Therefore\ldots ''

@@

\textcolor{blue}{We are intersted in the case of tolerant aneuploidy given that this represents a two step evolutionary rescue scenario. Change S11}

18. Line 438, "Decreasing the sensitive cell division rate will also decrease the cancer recurrence time." Note: This is a perfect example of how the leading word "Additionally" is unnecessary bc you use "also" in the sentence.

\textcolor{blue}{Change S12}

19. There a couple of places in the Discussion where you use ``fast enough'' to communicate the idea of timing with regard to aneuploidy. ``Fast enough'' indicates something about rate (and maybe timing), but ``early enough'' is about timing. Consider whether you want to use one or both to get this idea across. (Lines 474 and 477).

\textcolor{blue}{We have changed both usages to make it clear that we are talking specifically about the rate of occurrence of aneuploidy, not the timing. They now read ``as long as it occurs frequently enough'' and ``if the rate of occurrence of aneuploidy is high enough''.}

@@

\textcolor{blue}{Fixed.}

appendix D: Smith $\&$ Haigh 1974 should be Maynard Smith $\&$ Haigh 1974.

\textcolor{blue}{Fixed.}

%\textcolor{red}{Remus comments:}
%
%\textcolor{red}{We change "rapidly" to "easily" because $\tau_m<\tau_a$ conditioned on evolutionary rescue and the word "rapidly" suggests that aneuploid speeds up the process of rescue which is not true. Change RS1}
%
%\textcolor{red}{Recurrence time is a decreasing function of sensitive and aneuploid growth rates not an increasing function as stated in the manuscript. Change RS2}
%
%\textcolor{red}{"anuploids"$\rightarrow$"aneuploids". Change RS3}
%
%\textcolor{red}{"aneuplidies"$\rightarrow$"aneuploids". Change RS4}
%
%
%

### 2025-03-11 ms.tex
edited response to reviewer 1

Using multi-type branching processes, we derived the probability of evolutionary rescue of the tumor under the effects of aneuploidy, ranging from tolerance to partial resistance.
We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$. Notably, the latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued, as long as it is produced fast enough (\Cref{rescue_prob}A, \cref{eq:N_ratio}).

When aneuploid cells are partially resistant to the drug ($\Delta_s\ll0\ll\Delta_a\ll\Delta_m$), aneuploidy itself rescues the population (\Cref{rescue_threshold}A). 
When aneuploidy only provides tolerance to the drug ($\Delta_s\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
Instead, if the aneuploidy rate is fast enough (\cref{eq:N_ratio}) then it may act as a ``stepping stone'' through which the resistant mutant can appear more easily, given that the number of aneuploid cells declines slower than the number of drug-sensitive cells (\Cref{sampleTrajectories}).  %change RS1
In this scenario, aneuploidy provides two advantages. First, it delays the extinction of the population, providing more time for the appearance of a resistance mutation. %change S13
Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur. 
Together, this increases the cumulative number of mutants that arise (i.e., $N u v \lambda_s \lambda_a / \abs{\Delta_s \Delta_a}$). % change 39

>>>>

Using multi-type branching processes, we derived the probability of evolutionary rescue of the tumor under the effects of aneuploidy, ranging from tolerance to partial resistance.
We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
Notably, the latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued, as long as it occurs frequently enough (\Cref{rescue_prob}A, \cref{eq:N_ratio}).

When aneuploid cells are partially resistant to the drug ($\Delta_s\ll0\ll\Delta_a\ll\Delta_m$), aneuploidy itself rescues the population (\Cref{rescue_threshold}A). 
When aneuploidy only provides tolerance to the drug ($\Delta_s\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
Instead, if the rate of occurrence of aneuploidy is high enough (\cref{eq:N_ratio}) then aneuploidy may act as a ``stepping stone'' through which the resistant mutant can appear more easily, given that the number of aneuploid cells declines slower than the number of drug-sensitive cells (\Cref{sampleTrajectories}).  %change RS1
In this scenario, aneuploidy provides two advantages. First, it delays the extinction of the population, providing more time for the appearance of a resistance mutation. %change S13
Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur. 
Together, this increases the cumulative number of mutants that arise (i.e., $N u v \lambda_s \lambda_a / \abs{\Delta_s \Delta_a}$). % change 39
